---
title: "Machine Learning - Exercise 2"
date: "12.4.2022"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```


# 1. 
##### ISLR, Exercise 5.4.2, p. 219-220.
*We will now derive the probability that a given observation is part of a bootstrap sample. Suppose that we obtain a bootstrap sample from a set of n observations.*

(a) *What is the probability that the first bootstrap observation is not the jth observation from the original sample? Justify your answer.*

If we have n observations and draw a random draw from them, the probability to obtain the jth observation (a specific one) is 1/n. The complement of that is the probability to not obtain the jth observation. The complement is defined as $1 - 1/n = (n-1)/n$.

Give formulas and use correct wording (permutations)...










(b) *What is the probability that the second bootstrap observation is not the jth observation from the original sample?*

Bootstap samples are drawn with replacement, so the probabilities of each draw is the same. The probability is (n-1)/n


(c) *Argue that the probability that the jth observation is not in the bootstrap sample is (1 - 1/n)^n.*

If the probability that the jth observation is (n-1)/n for one specific draw, then we compute the probability that is not in any of the draws by multiplying. $(n-1)/n * (n-1)/n * ... * (n-1)/n$ (not in the first draw, and not in the second draw, ... , an not in the last draw). Bootstrap sampling uses the same sample size as the original sample (n). This means that we have ((n-1)/n)^n. (n-1)/n can also be written as 1 - 1/n, which means that we have (1-1/n)^n.


(d) *When n = 5, what is the probability that the jth observation is in the bootstrap sample?*

We use the complement
$1 - (1-1/n)^n = 1 - (1-1/5)^5 = 1 - (4/5)^5 = 0.67232$


(e) *When n = 100, what is the probability that the jth observation is in the bootstrap sample?*

$1 - (99/100)^{100} = 0.6339677$


(f) *When n = 10, 000, what is the probability that the jth observation is in the bootstrap sample?*

$1 - (9999/10000)^{10000} = 0.632139$
When n grows bigger, the probability is getting ever closer to 1 - 1/e


(g) *Create a plot that displays, for each integer value of n from 1 to 100, 000, the probability that the jth observation is in the bootstrap sample. Comment on what you observe.*

```{r 1g}

n = 1:100000
probs = 1 - ((n-1)/n)^{n}
plot(n, probs, log='xy')

```


(h) *We will now investigate numerically the probability that a bootstrap sample of size n = 100 contains the jth observation. Here j = 4. We repeatedly create bootstrap samples, and each time we record whether or not the fourth observation is contained in the bootstrap sample. Comment on the results obtained.*

```{r 1h}

set.seed(1)
store=rep(NA, 10000)
for(i in 1:10000){
  store[i]=sum(sample(1:100, rep=TRUE)==4)>0
}
mean(store) #0.6405

#Which is getting ever closer to 1 - 1/e
1-1/exp(1) #0.63212...

```
\pagebreak


# 2.
*Suppose that n = 10 and the observations are*
6.45, 1.28, -3.48, 2.44, -5.17, -1.67, -2.03, 3.58, 0.74, -2.14
*Write a script in R to simulate the fraction of the original observations not contained in a bootstrap sample. Use B = 10 000 bootstrap replications. Compare with the approximation 10/3.*

```{r 2}

set.seed(1)
n=10
obs = c(6.45, 1.28, -3.48, 
        2.44, -5.17, -1.67, 
        -2.03, 3.58, 0.74, -2.14)

store=rep(NA, 10000)
for(i in 1:10000){
  store[i]= (n - sum(obs %in% sample(obs, size=10, rep=TRUE)))/n
}
mean(store)


#If we compare the mean value of observations included in the bootstrap (instead of the fraction) it is
mean(store)*n

#Which is close to
10/3


```
\pagebreak


# 3 
##### ISLR, Exercise 8.4.2, p. 361
*It is mentioned in Section 8.2.3 that boosting using depth-one trees (or stumps) leads to an additive model: that is, a model of the form*

$$f(X) = \sum_{j=1}^p  f_j(X_j).$$

*Explain why this is the case. You can begin with (8.12) in Algorithm 8.2.*

\n
\n
Equation 8.12:
$\hat f(x) = \sum_{b=1}^B  \lambda \hat f^b(x)$


<!-- We are doing this summation for all bags -->
<!-- $\hat f(x) = \lambda  \sum_{b=1}^B (c_{1_b} \cdot I_{(X \in R_1)} + c_{2_b}   \cdot I_{(X \in R_2)})$ -->

<!-- Because each bag is split on one variable it leads to -->
<!-- $\hat f(x) = \lambda  \sum_{j=1}^p (c_{1_b} \cdot I_{(X \in R_1)} + c_{2_b}   \cdot I_{(X \in R_2)})$ -->



$$
\begin{equation}

\text{A stepwise solution:}\\
\text{A regular tree can be defined as}\\
f(X) = \sum_{m=1}^M c_m · 1_{(X \in R_m),}\\
\text{where M is the number of partitions, which for a stump is two}\\
\hat f^b(x) = c_{1_b} \cdot I_{(X \in R_1)} + c_{2_b}   \cdot I_{(X \in R_2).}\\

\text{We can insert this into equation 8.12}\\
\hat f(x) = \sum_{b=1}^B  \lambda \hat f^b(x) \\
=\sum_{b=1}^B \lambda(\alpha_b \cdot I(X_{j_b}<s_b) + \beta_b \cdot I(X_{j_b} \ge s_b)).\\

\lambda \text{ is a constant and can be moved outside the summation}\\
= \lambda \sum_{b=1}^B \alpha_b \cdot I(X_{j_b}<s_b) + \beta_b \cdot I(X_{j_b} \ge s_b).\\

\text{Because}\\
X_{j_1}, X_{j_2}, ..., X_{j_B} \in \{X_1, X_2, ..., X_p\},\\
\text{boosting using stumps leads to the additive model}\\
f(X) = \sum_{j=1}^p  f_j(X_j).


\end{equation}
$$

\pagebreak






# 4.
*Use the data USCompaniesdata.dta. Create a training set containing half of the observations, and a test set containing the remaining observations. Fit a tree with Return on Assets (roa_w) as the response and the other variables as predictors.*


```{r prep} 
library(haven) #For importing Stata data
library(tree) #For fitting trees
library(randomForest) #For bagging and randomforests
library(gbm) #For boosting

#Load Stata data with read_dta function from haven
USCompanies_data_winsorized = read_dta("USCompanies_data_winsorized.dta")
USData = subset(USCompanies_data_winsorized, select = -conm)
USData = na.omit(USData)

#Split into training and test set
set.seed(1)
train <- sample(1:nrow(USData), nrow(USData)/2)
test <- (-train)

# set.seed(1)
# train=sample(c(TRUE, FALSE), nrow(USData), rep=TRUE)
# test = !train

```


```{r 4}
#Fit a tree using function "tree" (in library "tree")
treeROA = tree(roa_w~. , USData, subset=train)
summary(treeROA)
#treeROA

#Plot
plot(treeROA); text(treeROA, pretty=0, cex=0.7, srt=25)

#MSE
yhat=predict(treeROA, newdata=USData[-train,])
ROAtest=USData[-train, "roa_w"]
plot(yhat, ROAtest$roa_w, ylab = "roa_w")
abline(0,1)
mean((yhat-ROAtest$roa_w)^2)

```
\pagebreak



# 5.
*Apply bagging to USCompaniesdata.dta. Compare the MSE of the tree in Exercise 4 with the MSE of the bagged trees.*
```{r 5}
#Bagging

#Bagging using the function randomForest,
#inside the randomForest library. When mtry is for all variables, it is bagging. 
bagROA = randomForest(roa_w~., USData, subset=train, mtry=ncol(USData)-1, importance =TRUE)
bagROA
plot(bagROA)

#MSE
yhat=predict(bagROA, newdata=USData[-train,])
ROAtest=USData[-train, "roa_w"]
plot(yhat, ROAtest$roa_w)
abline(0,1)
mean((yhat-ROAtest$roa_w)^2)

```
\pagebreak


# 6.
*Apply random forests to USCompaniesdata.dta. Does random forests provide an improvement over the bagged trees in Exercise 5?*
```{r 6}
#Random forests

set.seed(1)
#Fit a random forest using the function randomForest,
#inside the randomForest library. When mtry is less than
#than all variables in the data, it is a random forests model
forestROA = randomForest(roa_w~., USData, subset=train, importance = TRUE)

#Should be done on several different mtry
#sqrt(ncol(USData))
#ncol(USData)/3

forestROA
plot(forestROA)

importance(forestROA)
varImpPlot(forestROA)

#MSE
yhat=predict(forestROA, newdata=USData[-train,])
ROAtest=USData[-train, "roa_w"]
plot(yhat, ROAtest$roa_w)
abline(0,1)
mean((yhat-ROAtest$roa_w)^2)

```
\pagebreak



# 7. 
*Apply boosting to USCompaniesdata.dta. Which variables are the most important predictors in the boosted model?*
```{r 7}
#Boosting
set.seed(1)
#Boosting using the "gbm" function inside the "gbm" library
boostROA = gbm(roa_w~., USData[train,], distribution="gaussian", n.trees=500, interaction.depth=4, shrinkage = 0.1)
#all or only training set? cv.folds? check help(gbm)
#We use cross-validation to select B.


boostROA
summary(boostROA)
plot(boostROA, i="profit_margin_w"); plot(boostROA, i="ebitda_w")

#MSE
yhat=predict(boostROA, newdata=USData[-train,])
ROAtest=USData[-train, "roa_w"]
plot(yhat, ROAtest$roa_w)
abline(0,1)
mean((yhat-ROAtest$roa_w)^2)



```




